04-analisis

### 2024-06-12 01-introduction.qmd
revisada introduccion y antecedentes, temas menores formato
Update intro and analysis

---
author: "Equipo EDUMER"
format: html
editor: visual
bibliography: "input/bib/just-ineq-merit.bib"
csl: "input/bib/apa6.csl"
---

## Introduction

@@

In the present paper, we address the role of the perception of meritocracy on the justification of social inequalities by eighth-grade students in Chile, a country characterized by a highly stratified educational system. In particular, we focus in justifying inequalities in health, education and pensions, which are traditional social policy areas where acces to better services could be justified based on payment capacity, referred as to market justice [@lane_market_1986, @lindh_bringing_2023].  
Most of the research in this realm has only considered adults, leaving aside the study of how beliefs in this field develop at student age as well as the impact of the school context and the family as the main socialization agencies [@batruch_belief_2022]. Regarding schools, the way in which they deal with unequal conditions of origin has been linked to the *hidden curriculum* [@chafel_schooling_1997], whereby students learn about distributive norms in society and mechanisms of justification of social differences.

@@

++ párrafo contexto Chileno dar cifras/porcentajes de privatización de salud, educación y pensiones; asociar al tema de la calidad (lo público es peor); hay información sobre esto que se puede sacar del proyecto Fondecyt

>>>>

---
author: "Equipo EDUMER"
format: html
bibliography: "input/bib/just-ineq-merit.bib"
csl: "input/bib/apa6.csl"
---

# Introduction

@@

In the present paper, we address the role of the perception of meritocracy on the justification of social inequalities by eighth-grade students in Chile, a country characterized by a highly stratified educational system. In particular, we focus in justifying inequalities in health, education and pensions, which are traditional social policy areas where acces to better services could be justified based on payment capacity, referred as to market justice [@lane_market_1986, @lindh_bringing_2023].\
Most of the research in this realm has only considered adults, leaving aside the study of how beliefs in this field develop at student age as well as the impact of the school context and the family as the main socialization agencies [@batruch_belief_2022]. Regarding schools, the way in which they deal with unequal conditions of origin has been linked to the *hidden curriculum* [@chafel_schooling_1997], whereby students learn about distributive norms in society and mechanisms of justification of social differences.

@@

The Chilean case is particularly intriguing for studying market justice preferences. This country is characterized by acute and persistent economic inequality, which stands out in Latin America and among OECD countries [@flores_top_2020]. In Chile, the poorest 50% captures only 10% of the total income and has negative wealth, while the richest 1% receives almost 27% of the income and holds 49.6% of the wealth [@chancel_world_2022]. Much of this inequality is linked to the deep neoliberal reforms that institutionalized privatization and commodification of various economic sectors, including public services, introduced during the dictatorship (1973-1989) and expanded in democracy through concessions, demand credits, and specific regulatory frameworks [@ffrench-davis_reformas_2018]. This shift in economic policy has allowed the unprecedented emergence of markets in areas related to social reproduction, such as health, pensions, and education, with provision and access managed by private entities and segmented by individual payment capacity, heavily reliant on State subsidies [@boccardo_30_a]. In health, although the majority of the population uses the public insurance system (78.9%), 15.3% are served by private insurers (CASEN, 2022). The pension system is based on individual capitalization, with mandatory contributions managed by private administrators investing in the financial market, currently involving 11 million contributors (Superintendencia de Pensiones, 2024). In education, 30.6% of school enrollment is in public schools, 54.0% in State-subsidized private schools, and 9.3% in fully private schools, generally attracting higher-income groups (Mineduc, 2023).

### 2024-06-12 02-antecedentes.qmd
Update intro and analysis

theme_nice <- function() {
  theme_bw() +
    theme(text = element_text(family = "serif"))

}

theme_set(theme_nice())
```

@@

elsoc_2016 <- elsoc_2016 %>% 
  dplyr::select(Pensions = d02_01, 
                Education = d02_02, 
                Health = d02_03) %>% 
  mutate_all(.funs = ~ set_na(., na = c(-666,-777,-888,-999), drop.levels = F) %>% 
               sjmisc::rec(., rec = "rev", append = F) %>% 
               factor(.,
                      levels = c(1,2,3,4,5),
                      labels = c("Strongly agree",
                                 "Agree",
                                 "Neither agree nor disagree",
                                 "Desagree",
                                 "Strongly desagree"))) %>% 
  na.omit()

elsoc_2016 %>% 
  pivot_longer(cols = everything(),
               names_to = "justice", 
               values_to = "value") %>% 
  mutate(justice = factor(justice, 
                          levels = c("Health", "Pensions", "Education"))) %>% 
  group_by(justice, value) %>%
  summarise(total = n()) %>% 
  mutate(prop = prop.table(total)*100) %>% 
  ggplot(aes(x = value, y = prop, fill = value)) +
  geom_col(position = "dodge", width = .9, alpha = 1, show.legend = T) +
  scale_y_continuous(labels = function(prop){paste0(prop, "%")},
                     limits = c(0,55),
                     n.breaks = 7,
                     expand = c(0,1)) +
  facet_wrap(.~justice, scales = "fixed") +
  geom_text(mapping = aes(label = scales::percent(prop, accuracy = 0.1, scale = 1, decimal.mark = ".")),
            vjust = -0.5,
            size = 3,
            position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c( "#F1EEF6", "#BDC9E1", "#b3b3b3ff", "#74A9CF", "#0570B0")) +
  labs(x = NULL,
       y = "Proportion (%)",
       fill = NULL,
       caption = paste0("Source: Authors calculation based on ELSOC 2016",
                        " (N = ",dim(elsoc_2016)[1],")")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 0))

@@

The original definition of merit is a combination of effort and talent [@young_rise_1958], and a meritocracy is a distributive system where merit is the main criteria for allocating valuable goods and rewards. From a sociological perspective, meritocracy has been used in research on social mobility to characterize societies with low mobility that threaten the meritocratic ideal [@goldthorpe_myth_2003]. More recently, sociology and social psychology research has attended to subjective aspects related to the support for meritocratic principles in different societies, such as beliefs in meritocracy [@castillo_multidimensional_2023a; @mijs_unfulfillable_2016; @mijs_paradox_2019]. The label of beliefs in this realm covers different concepts, but in this paper we refer to a conceptual model developed by @castillo_multidimensional_2023 which identify two main aspects: preferences and perceptions for meritocracy.

>>>>

theme_set(theme_bw())
```

@@

elsoc_2016 <- elsoc_2016 %>% 
  dplyr::select(Pensions = d02_01, 
                Education = d02_02, 
                Health = d02_03) %>% 
  mutate_all(.funs = ~ set_na(., na = c(-666,-777,-888,-999), drop.levels = F) %>% 
               factor(.,
                      levels = c(1,2,3,4,5),
                      labels = c("Strongly desagree",
                                 "Desagree",
                                 "Neither agree nor disagree",
                                 "Agree",
                                 "Strongly agree"))) %>% 
  na.omit()

elsoc_2016 %>% 
  pivot_longer(cols = everything(),
               names_to = "justice", 
               values_to = "value") %>% 
  mutate(justice = factor(justice, 
                          levels = c("Health", "Pensions", "Education"))) %>% 
  group_by(justice, value) %>%
  summarise(total = n()) %>% 
  mutate(prop = prop.table(total)*100) %>% 
  ggplot(aes(x = value, y = prop, fill = value)) +
  geom_col(position = "dodge", width = .9, alpha = 1, show.legend = T) +
  scale_y_continuous(labels = function(prop){paste0(prop, "%")},
                     limits = c(0,55),
                     n.breaks = 7,
                     expand = c(0,1)) +
  facet_wrap(.~justice, scales = "fixed") +
  geom_text(mapping = aes(label = scales::percent(prop, accuracy = 0.1, scale = 1, decimal.mark = ".")),
            vjust = -0.5,
            size = 3,
            position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("#F1EEF6", "#BDC9E1", "#b3b3b3ff", "#74A9CF", "#0570B0")) +
  labs(x = NULL,
       y = "Proportion (%)",
       fill = NULL,
       caption = paste0("Source: Authors calculation based on ELSOC 2016",
                        " (N = ",dim(elsoc_2016)[1],")")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 0))

@@

The original definition of merit is a combination of effort and talent [@young_rise_1958], and a meritocracy is a distributive system where merit is the main criteria for allocating valuable goods and rewards. From a sociological perspective, meritocracy has been used in research on social mobility to characterize societies with low mobility that threaten the meritocratic ideal [@goldthorpe_myth_2003]. More recently, sociology and social psychology research has attended to subjective aspects related to the support for meritocratic principles in different societies, such as beliefs in meritocracy [@castillo_multidimensional_2023; @mijs_unfulfillable_2016; @mijs_paradox_2019]. The label of beliefs in this realm covers different concepts, but in this paper we refer to a conceptual model developed by @castillo_multidimensional_2023 which identify two main aspects: preferences and perceptions for meritocracy.

### 2024-06-12 04-analisis.qmd
Update intro and analysis

#windowsFonts(`Times New Roman` = windowsFont("Times New Roman"))

theme_nice <- function() {
  theme_bw() +
    theme(text = element_text(family = "serif"))

}

theme_set(theme_nice())

@@

ggarrange(plot_soc_effort, plot_school_effort, plot_soc_talent,
          plot_school_talent, plot_deservingness, leg, ncol = 2, nrow = 3)
```

@@

The results of the cumulative link mixed models for the ordinal dependent variables of justice in differential access to health, pensions and education by income are presented in @tbl-ordinal-reg[^1].

@@

```{r echo=FALSE, results='asis'}
#| tbl-cap: "Cumulative link multilevel models of differential acces justificacion of education, health and pensions"
#| label: tbl-ordinal-reg

@@

@tbl-ordinal-reg demonstrates a similar trend for the three dependent variables of justice in differential access to pensions, education, and health, with all effects being statistically significant. In the context of school meritocracy, the effects are mixed: as perceived school talent increases, the justification for differentiated access to these services increases; conversely, as perceived school effort increases, the justification decreases, holding all other variables constant. Regarding the variables of social meritocracy, the three variables of talent, effort, and deservingness show that as these increase, the justification for differentiated access to pensions, education, and health also increases. Overall, these results provide partial support for our first hypothesis (H1a), which posited that students who perceive higher levels of meritocracy in school would be more likely to justify differential access to pensions, health, and education. This is evidenced by the positive effect of perceiving talent as important for achieving good grades. However, the results strongly support our second hypothesis (H1b), as a higher perception of societal meritocracy among students tends to increase the justification for market-based justice in these social services. CAMBIAR

Contextual effects for ordinal models.

@@

```{r echo=FALSE, results='asis'}
#| tbl-cap: "Multilevel models for meritocracy perception and market justice preferences"
#| label: tbl-individual-reg

ccoef <- list(
  "(Intercept)" = "Intercept",
  inteligencia_esc = "School talent",
  esfuerzo_esc = "School effort",
  inteligencia_soc = "Social talent",
  esfuerzo_soc = "Social effort",
  merito_soc = "Deservingness",
  "educacion_recEd Media" = "Secondary",
  "educacion_recEd Tecnica" = "Higher tec.",
  "educacion_recUniversidad o posgrado" = "University or posgraduate",
  "educacion_recNs/Nr" = "Missing",
  "libros_recMás de 25" = "More than 25 books (Ref.= Less than 25)",
  "cod_grupoMedio bajo" = "SES Medium low",
  cod_grupoMedio = "SES Medium",
  "cod_grupoMedio alto" = "SES Medium high",
  cod_grupoAlto = "SES High",
  simceMedio = "Simce Medium",
  simceAlto = "Simce High"
)

@@

[^1]: This table shows the effects of individual and contextual-level independent variables related to our hypothesis; however, the models also incorporate the control variables utilized in the study for each of the three dependent variables separately (complete models are available in the [Appendix](#appendix)).

>>>>

theme_set(theme_bw())

@@

ggarrange(plot_soc_effort, plot_school_effort, plot_soc_talent,
          plot_school_talent, plot_deservingness, leg, ncol = 2, nrow = 3) +
  labs(caption = paste0("Source: Authors calculation based on First Study of Civic Education 2017",
                             " (N = ",dim(data_rec)[1],")"))
```

@@

This section presents the results of the multilevel models. We begin by showing the results of the cumulative link mixed models in @tbl-ordinal-reg for the ordinal dependent variables of justice in health, pensions, and education separately, to determine if the effect of the predictors differs across these items. Then, we present the results of the linear mixed-effects models for the average index of preference for market justice in @tbl-lineal-reg[^1]. For the health item, the intraclass correlation (ICC) obtained for the null model (not shown) indicates that 5% of the total variance of this item is associated with specific characteristics of the schools, while it is 3% for the pensions item and 4% for the education item (ICC based on [@hox_multilevel_2010, p.15]). For the average index of market justice preferences, the ICC reaches 4%, indicating that overall, only a small percentage of the total variance in students’ market justice preferences is associated with school characteristics, thus limiting the possibility of finding effects at this second level of analysis.

[^1]: The tables shows the effects of individual and contextual-level independent variables related to our hypothesis; however, the models also incorporate the control variables utilized in the study for each of the three dependent variables separately and for the average index (complete models are available in the [Appendix](#appendix)).

@@

```{r echo=FALSE, results='asis'}
#| tbl-cap: "Cumulative link multilevel models of differential acces justificacion of health, pensions, and education"
#| label: tbl-ordinal-reg

@@

Model 1 in @tbl-ordinal-reg for the three dependent variables of justice in differential access to health, pensions, and education demonstrates a similar trend regarding the effect of the perception of meritocracy on these items, with all effects being statistically significant (*p* < 0.05). Regarding the variables of social meritocracy, the three variables—talent, effort, and deservingness—show that as these increase, the justification for differentiated access to pensions, education, and health also increases. In the context of school meritocracy, the effects are mixed: as perceived school talent increases, the justification for differentiated access to these services increases; conversely, as perceived school effort increases, the justification decreases, holding all other variables constant.

Additionally, Model 1 shows the effect of family status, measured through the highest educational level of the parents and the number of books at home. For the three items—health, pensions, and education—a negative but statistically non-significant association with parents' educational level is observed. However, the results suggest a negative and statistically significant effect of the number of books at home for market justice in health and pensions, indicating that those with more than 25 books at home tend to justify less differentiated access in these areas compared to those with fewer than 25 books, controlling for other variables. This association is not observed for market justice in education.

Model 2 in @tbl-ordinal-reg incorporates school-level variables for the three items of market justice in health, pensions, and education. Regarding the socioeconomic level of the schools, the results show that high socioeconomic status schools maintain greater preferences for market justice in education compared to low socioeconomic status schools, with a statistically significant association (*p* < 0.05). This effect is not observed for the health and pensions variables. Moreover, the academic performance of schools shows negative and statistically significant effects for the three items of market justice separately. The findings suggest that schools with higher academic performance tend to justify market allocation principles less for access to health, pensions, and education compared to low-performing schools. This difference is also observed in schools with medium performance, but only for the domains of health and pensions.

@@

```{r echo=FALSE, results='asis'}
#| tbl-cap: "Linear mixed-effects models for meritocracy perception and market justice preferences"
#| label: tbl-lineal-reg

ccoef <- list(
  "(Intercept)" = "Constant",
  inteligencia_esc = "School talent",
  esfuerzo_esc = "School effort",
  inteligencia_soc = "Social talent",
  esfuerzo_soc = "Social effort",
  merito_soc = "Deservingness",
  "educacion_recEd Media" = "Secondary",
  "educacion_recEd Tecnica" = "Higher tec.",
  "educacion_recUniversidad o posgrado" = "University or posgraduate",
  "educacion_recNs/Nr" = "Missing",
  "libros_recMás de 25" = "More than 25 books (Ref.= Less than 25)",
  "cod_grupoMedio bajo" = "SES Medium low",
  cod_grupoMedio = "SES Medium",
  "cod_grupoMedio alto" = "SES Medium high",
  cod_grupoAlto = "SES High",
  simceMedio = "Simce Medium",
  simceAlto = "Simce High"
)

@@
